---
title: "HW1"
author: "Aram Abrahamyan"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Part 1: Data Cleaning and Exploration
```{r}
library(dplyr)
library(lubridate)
```

#### 1. Load the dataset. Check the first 5 rows.
```{r}
df = read.csv('crime_data.csv')
head(df)
```
#### 2. Identify columns with missing values and their respective counts. Drop columns where more than 50% of the data is missing (store this version as a new dataset).
```{r}
c = colSums(is.na(df))
c
```
```{r}
df_cl = df[, colSums(is.na(df)) <= 0.5 * nrow(df)]
df_cl
```

#### 3. Convert the DATE OCC column to a datetime format. Extract the year, month, and day into separate columns. Create a new column for the hour using the TIME OCC column.
```{r}
df_cl <- df_cl %>% mutate(
DATE.OCC = as.Date(DATE.OCC, format = "%m/%d/%Y"),
Year = format(DATE.OCC, "%Y"),
Month = format(DATE.OCC, "%m"),
Day = format(DATE.OCC, "%d"),
Hour = as.integer(TIME.OCC / 100)
)
head(df_cl, 5)
```
#### 4. Filter the dataset for crimes that occurred in 2023. Further filter crimes with the description BURGLARY in the Crm Cd Desc column.
```{r}
df_1 = df_cl %>% filter(Year == 2023, Crm.Cd.Desc == "BURGLARY")
head(df_1, 5)
```
#### 5. Group the data by AREA NAME and calculate the total number of crimes and the average victim age. Sort the results by total crimes in descending order.
```{r}
df_2 = df_cl %>% group_by(AREA.NAME) %>% summarise(
  total_crimes = n(),
  avg_vctm_age = mean(Vict.Age, na.rm=TRUE)
) %>% arrange(desc(total_crimes))
head(df_2, 5)
```
### Part 3: Further Exploration (R only)
#### 1. Group the data by Month and count the number of crimes.
```{r}
df_3 = df_cl %>% group_by(Month) %>%
  summarise(total_crimes = n())
head(df_3, 5)
```
#### 2. Count the number of crimes where a weapon was used (Weapon Used Cd is not null).
```{r}
df_4 = df %>% filter(!is.na(Weapon.Used.Cd)) %>%
  summarise(total_crimes = n())
df_4$total_crimes
```
#### 3. Group the data by Premis Desc and count the number of crimes.
```{r}
df_5 = df_cl %>% group_by(Premis.Desc) %>%
  summarise(total_crimes = n())
head(df_5, 5)
```
### Part 4: Advanced Analysis
#### 1. Create a new column, Severity Score, based on the following rules:
#### • Assign 5 points if a weapon was used (Weapon Used Cd is not null).
#### • Assign 3 points for crimes under BURGLARY.
#### • Assign 1 point for all other crimes.
#### • Group by AREA NAME and find the total severity score for each area.
```{r}
df <- df %>%
  mutate(Severity.Score = 
    ifelse(!is.na(Weapon.Used.Cd), 5, 0) +
    ifelse(Crm.Cd.Desc == "BURGLARY", 3, 0) +
    ifelse(is.na(Weapon.Used.Cd) & Crm.Cd.Desc != "BURGLARY", 1, 0)
  )
df
```

